---
title: "ROI Value-at-Risk Modeling for Commerical Real Estate Investments"
subtitle: "Generating Sensitivity Analyses using Monte Carlo Simulation"
output: github_document
---

This is a basic model and simulation exercise simulating how forecasting and Monte Carlo simulation can be used to create Value-at-Risk sensitivity analyses. Here, we consider a hypothetical Manhattan Office building purchased in year T and sold in year T+10. By roughly estimating and then forcasting the Net Operating Income and the Exit Cap Rate, we are able to determine possible ROI and subsequently Value-at-Risk.


```{r setup, include=TRUE, message = FALSE, warning=FALSE}
rm(list=ls())
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

library(tidyverse)
library(stringr)
library(forecast)
library(zoo)


set.seed(608)
```

# User Inputs

```{r}
# a fictitious purchase price
purchase_price <- 31500000

# years to hold the property
year_hold <- 10

# current year NOI
current_NOI <- 3500000

# input the number simulations to run
n_sims <- 10000

```

# Exit Caps

```{r}
# Exit Cap data from RCA
RCA_trxns <- read_csv("data/RCA_manhat_office_trxs.csv")

RCA_trxns <- RCA_trxns %>% mutate(Date = as.Date(Date, format = "%m/%d/%y"))

manhat_office_caps <- 
  RCA_trxns %>% select(`Cap Rate`, Date) %>% filter(!is.na(`Cap Rate`)) %>% 
  mutate(Year = lubridate::year(Date)
         , Month = lubridate::month(Date)
         , YearMonth = as.Date(paste(Year,Month,"01",sep="-"), format = "%Y-%m-%d"))

head(manhat_office_caps)

```

## Forecasting exit cap rates out 10 years

```{r}
cap_rate_means <- 
  manhat_office_caps %>% 
  group_by(YearMonth) %>% 
  summarise(Mean_cap = mean(`Cap Rate`, na.rm = T)
           , SD_cap = sd(`Cap Rate`, na.rm = T)
           , count = n()
           ) %>% 
  mutate(SD_cap = zoo::na.locf(SD_cap, na.rm = T))

# visualize exit cap rates and standard deviations over time (Office, Manhattan)
cap_rate_means %>% 
  mutate(upper_sd = Mean_cap+SD_cap, lower_sd = Mean_cap - SD_cap) %>% 
  ggplot()+
  aes(x = YearMonth, y = Mean_cap)+
  geom_ribbon(aes(ymin = lower_sd, ymax = upper_sd), alpha = 0.5, fill = "skyblue")+
  geom_line(size = 2, color = "black")
```


```{r}
# forecasting exit cap rates 10 years
cap_rate_means_ts <- ts(cap_rate_means$Mean_cap, start = c(2007,7), end = c(2017,7), frequency = 12)
mod <- cap_rate_means_ts %>% smooth() %>% auto.arima()
cap_for <- forecast(cap_rate_means_ts, model = mod, h = 12*year_hold)
plot(cap_for)
```


```{r}
# extract the point forecasts into a dataframe
# 95% CI is 2 standard deviations, so, 95% CI/2  = one SD
date_seq <- seq.Date(from = as.Date("08-2017-01", format = "%m-%Y-%d")
                     , to = as.Date("07-2027-01", format = "%m-%Y-%d")
                     , by= "month")

cap_forc_data <- 
  tibble("YearMonth" = date_seq
         , "Mean_cap" = as.numeric(cap_for$mean)
         , "SD_cap" = as.numeric(cap_for$upper[,2]-as.numeric(cap_for$mean))/2
         ) %>% 
  mutate("Lower_bound" = Mean_cap - SD_cap, "Upper_bound" = Mean_cap + SD_cap) 

# visualize the forecasts + SD
cap_forc_data %>% 
  ggplot()+
  aes(x = YearMonth, y = Mean_cap, ymin = Lower_bound, ymax = Upper_bound)+
  geom_ribbon(alpha = 0.5, fill = "skyblue")+
  geom_line(size = 2, color = "black")+
  theme_minimal()

```


# Forecasting Rent Growth

```{r}
# Office rent data from CoStar
costar_manhat_office <- read_csv("data/Costar_office_manhat_stats.csv")

office_rents <- 
  costar_manhat_office %>% 
  filter(Quarter!="QTD") %>% 
  mutate(Year = str_replace(Quarter," Q[0-9]","")
         , Quart = str_extract(Quarter,"Q[0-9]")
         , YearQuarter = lubridate::yq(paste(Year,Quart))
         ) %>% 
  select(YearQuarter,`Office Gross Rent Overall`) %>% 
  arrange(YearQuarter) %>% 
  mutate(QoQ = (`Office Gross Rent Overall`-lag(`Office Gross Rent Overall`,1))/lag(`Office Gross Rent Overall`,1)) %>% 
  mutate(QoQ = if_else(is.na(QoQ),0,QoQ)) %>% 
  mutate(Cumlative_RG = cumsum(QoQ))

head(office_rents)

```



## Visualizig office rent growth

```{r}
office_rents %>% 
  ggplot()+
  aes(x = YearQuarter, y = QoQ)+
  geom_col()+
  geom_line(aes(y = Cumlative_RG))+
  theme_minimal()+
  labs(title = "Manhattan Office Quarterly Rent Growth"
       , y = NULL
       , x = NULL)
```


## Forecasting cumulative rent grwoth over 10 years

```{r}

rent_growth_means_ts <- ts(office_rents$Cumlative_RG, start = c(1996,3), end = c(2017,2), frequency = 4)

fit <- stl(rent_growth_means_ts, s.window="periodic", t.window = 4, l.window = 10)
fcast <- forecast(fit, method="naive", h = 4*year_hold)
plot(fcast)
```

```{r}
# extract the point forecasts into a dataframe
# 95% CI is 2 standard deviations, so, 95% CI/2  = one SD
date_seq2 <- seq.Date(from = as.Date("07-2017-01", format = "%m-%Y-%d"), to = as.Date("04-2027-01", format = "%m-%Y-%d"), by= "quarter")
rent_forc_data <- 
  tibble("YearMonth" = date_seq2
         , "Mean_rent" = as.numeric(fcast$mean)
         , "SD_rent" = (fcast$upper[,2]-fcast$mean)/2
         ) %>% 
  mutate("Lower_bound" = Mean_rent - SD_rent, "Upper_bound" = Mean_rent + SD_rent) 


# adjust rent growth data so that it is growth relative to today
set.seed(2017)
rent_forc_data <- 
  rent_forc_data %>% 
  mutate(rent_adj = (row_number()*0.5)*rnorm(n = nrow(rent_forc_data), mean = 0.5, sd = 0.2)) %>% 
  mutate(rent_adj = rnorm(n = nrow(rent_forc_data), mean = 0.5, sd = 0.3)) %>% 
  mutate(Mean_rent = Mean_rent + rent_adj, SD_rent = SD_rent + rent_adj, Lower_bound = Lower_bound + rent_adj, Upper_bound = Upper_bound + rent_adj) %>% 
  mutate_at(vars(Mean_rent:Upper_bound), .funs = function(x) x - (office_rents$Cumlative_RG[which(office_rents$YearQuarter==max(office_rents$YearQuarter))])) 

# visualize the forecasts + SD
rent_forc_data %>% 
  ggplot()+
  aes(x = YearMonth, y = Mean_rent, ymin = Lower_bound, ymax = Upper_bound)+
  geom_ribbon(alpha = 0.5, fill = "skyblue")+
  geom_line(size = 2, color = "black")+
  theme_minimal()

```


# Generating simulations

```{r}

# isolate the exit year:
exit_year_data_cap <- cap_forc_data %>% filter(row_number()==year_hold*12) #monthly
exit_year_data_rents <- rent_forc_data %>% filter(row_number()==year_hold*4) #quarterly

```

## Simulation of cap rates
```{r}

exit_caps_sim <- abs(rnorm(n = n_sims, mean = exit_year_data_cap$Mean_cap, sd = exit_year_data_cap$SD_cap))

exit_caps_sim %>% 
  as_data_frame() %>% 
  mutate(buckets = cut(value,breaks = seq(0,max(.),by=0.005))) %>% 
  group_by(buckets) %>% 
  summarise(count = n()) %>% 
  filter(!is.na(buckets)) %>% 
  mutate(probability = count/sum(count)) %>% 
  ggplot()+
  aes(x = buckets, y = probability)+
  geom_col()+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(labels = scales::percent)+
  labs(title = "Simulated Proability of Exit Cap Rates"
       , x = "Exit Cap Rates")
```


## Simulation of rent growth:
```{r}
exit_year_rents_sim <- rnorm(n = n_sims, mean = exit_year_data_rents$Mean_rent, sd = abs(exit_year_data_rents$SD_rent))
```

```{r}

exit_year_rents_sim %>% 
  as_data_frame() %>% 
  filter(is.finite(value)) %>% 
  mutate(buckets = cut(value
                       ,breaks = seq(from = 0, to = range(exit_year_rents_sim)[2],by=0.05)
                       )
         ) %>% 
  group_by(buckets) %>% 
  summarise(count = n()) %>% 
  filter(!is.na(buckets)) %>% 
  mutate(probability = count/sum(count)) %>% 
  ggplot()+
  aes(x = buckets, y = probability)+
  geom_col()+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(labels = scales::percent)+
  labs(title = "Simulated Proability of Rent Growth Rate"
       , x = "Rent Growth")
  
  
```


## Simulation of NOI:
```{r}
exit_year_noi_sim <- (exit_year_rents_sim*current_NOI)+current_NOI

exit_year_noi_sim  %>% 
  as_data_frame() %>% 
  mutate(buckets = cut(value,breaks = seq(0,max(.),by=250000)
                       )
         ) %>% 
    group_by(buckets) %>% 
    summarise(count = n()) %>% 
    filter(!is.na(buckets)) %>% 
    mutate(probability = count/sum(count)) %>% 
    ggplot()+
    aes(x = buckets, y = probability)+
    geom_col()+
    theme_minimal()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    scale_y_continuous(labels = scales::percent)+
    labs(title = "Simulated Proability of Exit Year NOI"
         , x = "NOI ($ Millions)")
```

## Simulation of sales price
```{r}
sale_price_sim <- exit_year_noi_sim/exit_caps_sim

summary(sale_price_sim)

sale_price_sim %>% 
 as_data_frame() %>% 
  mutate(buckets = cut(value, breaks = seq(0,max(.),by=10000000))) %>% 
    group_by(buckets) %>% 
    summarise(count = n()) %>% 
    filter(!is.na(buckets)) %>% 
    mutate(probability = count/sum(count)) %>% 
    ggplot()+
    aes(x = buckets, y = probability)+
    geom_col()+
    theme_minimal()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    scale_y_continuous(labels = scales::percent)+
    labs(title = "Simulated Proability of Exit Year Sales Price"
         , x = "Exit Sale Price")

```

# Simualtion of NOI
```{r}

ROI <- round((sale_price_sim - purchase_price)/purchase_price,2)

ROI %>% 
  as_data_frame() %>% 
  mutate(buckets = round(value,1)) %>% 
  group_by(buckets) %>% 
  summarise(count = n()) %>% 
  filter(!is.na(buckets)) %>% 
  mutate(probability = count/sum(count)) %>% 
  ggplot()+
  aes(x = buckets, y = probability)+
  geom_col()+
  theme_minimal()+
  scale_x_continuous(breaks = seq(0,6,by=1), labels = scales::percent)+
  scale_y_continuous(labels = scales::percent)+
  labs(title = "Simulated Proability of Exit Year ROI"
       , x = "10 Year ROI")

```

Of course, it wouldn't take much additional work to calculate probable IRR's or other investment metrics. 

